---
title: Calling Rust Functions from R using extendr - II
author: jModeler
date: '2025-05-03'
slug: []
categories: []
tags:
  - Rust
  - R
authors: []
description: ''
externalLink: ''
series: []
---

In a [previous post](https://jmodeler.github.io/posts/2025-04-27-calling-rust-functions-from-r-using-extendr/), I called a Rust function from R and did a speed test of that function against an equivalent R function. At the end of the post, I mentioned that we could use [memoization](https://en.wikipedia.org/wiki/Memoization) to speed up the R function. I proceed to do that here.

The speed test in the previous post might have been unfair to R, since it is well known that [R is not good with recursive function calls](https://stackoverflow.com/questions/6807068/why-is-my-recursive-function-so-slow-in-r). I now create a memoized version of the R function here:

```{r, echo=TRUE, eval=TRUE}
library(rextendr)
library(ggplot2)
library(microbenchmark)

# write memoized r function
fibonacci <- function(n) {
  fib <- vector('numeric', n)
  for(ii in 1:(n+1)) { 
    # R vector indices start at 1
    if(ii==1) {
      fib[ii] <- 0
    } else if(ii==2) {
      fib[ii] <- 1
    } else {
      fib[ii] <- fib[ii-1] + fib[ii-2]
    }
  }
  return(fib[n+1])
}

# run the R function
r_fn_result = fibonacci(15)

print(r_fn_result)

# test run time for a larger term number
system.time(fibonacci(40))
```

Note that the function returns the correct value for the `\(15^{th}\)` term and is much faster now for generating the value of the 40th Fibonacci term (it's even faster than the Rust function, which uses recursive function calls -- [see the previous post](https://jmodeler.github.io/posts/2025-04-27-calling-rust-functions-from-r-using-extendr/))

I now write a memoized version of the Rust function, and call it in R:
```{r, echo=TRUE, eval=TRUE}
# write memoized Rust function
rust_memoized_code <- "
#[extendr]
fn fibonacci_rust(n:usize) -> i64 {
  let mut fibn: Vec<i64> = vec![0;(n+1)];
  for ii in 0..=n {
    if ii==0 {
    fibn[ii] = 0;
  } else if ii==1 {
    fibn[ii] = 1;
  } else {
    fibn[ii] = fibn[ii-1] + fibn[ii-2];
  }
  }
  return fibn[n];
}"

# compile source code
rust_source(code = rust_memoized_code, quiet=TRUE)

# run the rust function
rustm_fn_result = fibonacci_rust(15)

print(rustm_fn_result)
```

Which, again is the correct value for the `\(15^{th}\)` term. I now do a microbenchmark test:

```{r, eval=TRUE, echo=TRUE}
# use microbenchmark to compare runtimes
# list of test values
values <- c(40, 50, 60)
# list to store microbenchmark results
compare <- vector('list', length(values))
for (ii in 1:length(values)) {
  compare[[ii]] <- microbenchmark(fibonacci(values[ii]), fibonacci_rust(values[ii]), times = 1000)
  # Change labels for plotting convenience later
  # Convert expressions to character strings
  expr_char <- as.character(compare[[ii]]$expr)
  
  # Map to new labels
  expr_char[expr_char == "fibonacci(values[ii])"] <- "R"
  expr_char[expr_char == "fibonacci_rust(values[ii])"] <- "Rust"
  
  # Assign the new factor with correct levels
  compare[[ii]]$expr <- factor(expr_char, levels = c("R", "Rust"))
}

# plot benchmark times, term number = 40
autoplot(compare[[1]]) +
  ggtitle(sprintf("Rust vs R: Microbenchmark Results for Term Number: %d", values[1]))
# plot benchmark times, term number = 50
autoplot(compare[[2]]) +
  ggtitle(sprintf("Rust vs R: Microbenchmark Results for Term Number: %d", values[2]))
# plot benchmark times, term number = 60
autoplot(compare[[3]]) +
  ggtitle(sprintf("Rust vs R: Microbenchmark Results for Term Number: %d", values[3]))

```

Note that while the R function appears to be faster when _n=40_, the Rust function tends to do better speed wise when _n_ gets larger. That said, the Rust function appears to have instances where it runs much longer than the R function.

