---
title: Calling Rust Functions from R using extendr
author: jModeler
date: '2025-04-27'
slug: []
categories: []
tags:
  - Rust
  - R
authors: []
description: ''
externalLink: ''
series: []
---
In this post, I attempt to call a [Rust](https://www.rust-lang.org/) function from an [R](https://cran.r-project.org/) session and compare the runtime of that function to a similar function written in R. 

The [rextendr](https://github.com/extendr/rextendr) R library allows us to call rust functions from an R session. As the repository linked above mentions, you "[need to setup a working Rust toolchain](https://cran.r-project.org/)". 

I write a Rust function to get the `\(n^{th}\)` [Fibonacci term](https://en.wikipedia.org/wiki/Fibonacci_sequence), and then compile it using the `rust_source()` function from the `rextendr` library, as shown below:

```{r, eval=FALSE, echo=TRUE}
library(rextendr)
library(parallel)

# write code to calculate the nth fibonacci term
rust_code <- "
#[extendr]
fn fibonacci_rust(n:i32) -> i64 {
  let mut fibn:i64 = 0;
  if n==0 {
    return 0;
  } else if n==1 {
    return 1;
  } else {
    return fibonacci_rust(n-1) + fibonacci_rust(n-2);
  }
}
"

# compile source code
rust_source(code = rust_code)

# run the rust function
rust_fn_result = fibonacci_rust(15)

# print the value 
print(rust_fn_result)
```

```{r, eval=TRUE, echo=FALSE}
load('code/results.RData')
print(rust_fn_result)
```

Which is the correct value of the `\(15^{th}\)` Fibonacci number. Note that the Rust code begins with the `#[extendr]` preamble, otherwise the function `fibonacci_rust()` will not be available in the R environment. 
I now create a similar function in R:

```{r, eval=FALSE, echo=TRUE}
# write r function
fibonacci <- function(n) {
  if(n==0) {
    return(0)
  } else if(n==1) {
    return(1)
  } else {
    return(fibonacci(n-1) + fibonacci(n-2))
  }
}

# run the R function
r_fn_result = fibonacci(15)

# print the value 
print(r_fn_result)
```

```{r, eval=TRUE, echo=FALSE}
load('code/results.RData')
print(r_fn_result)
```

Which is the correct value again. I now test the runtime of each function for a series of values starting from 1 and ending at 40. In addition, I also give Rust a handicap: I'll run the R function in parallel, while Rust will run serially.
```{r, eval=FALSE, echo=TRUE}
# set input values
n <- 1:40

# create functions that collect elapsed times
elapsed_time_rust <- function(x) {
  system.time(fibonacci_rust(x))['elapsed']
}

elapsed_time_r <- function(x) {
  system.time(fibonacci(x))['elapsed']
}

# call rust function, also note the time it takes to run for all inputs
rust_total_runtime <- system.time({
  rust_times <- sapply(n, FUN = elapsed_time_rust) 
})

# running the R function in parallel, since it will take longer as n gets larger
num_cores <- max(1, detectCores() - 1)

print(num_cores)

# Create a cluster with the specified number of worker processes
cl <- makeCluster(num_cores)

# Export the helper function to all workers
clusterExport(cl, varlist = c("fibonacci"))

# Use parLapply to apply the function in parallel
r_total_runtime <- system.time({
  r_times <- parLapply(cl, n, fun = elapsed_time_r)
})

# Simplify the result into a vector
r_times <- unlist(r_times)

# stop cluster
stopCluster(cl)

# print times taken
print("Time taken to calculate all fibonacci terms by Rust (serially):")
print(rust_total_runtime)
print("Time taken to calculate all fibonacci terms by R (in parallel):")
print(r_total_runtime)
```

```{r, echo=FALSE, eval=TRUE}
print("Time taken to calculate all fibonacci terms by Rust (serially):")
print(rust_total_runtime)
print("Time taken to calculate all fibonacci terms by R (in parallel):")
print(r_total_runtime)
```

**Rust takes 6 seconds to get the first 40 fibonacci terms in serial mode, while R takes 250 seconds (roughly 4 minutes) to do the same thing, _in parallel_!** (Note: I used 11 cores on my machine for parallel execution).

Let's also plot the time taken to calculate each term:
```{r, echo=TRUE, eval=TRUE}
plot(n, rust_times, type = "l", col = "#D84B16", lwd = 2, xlab = "Fibonacci Term Number", 
     ylab = "Time Taken (Seconds)", main = "Rust vs R Time Comparison",
     ylim = c(0, max(rust_times, r_times) + 1))

lines(n, r_times, col = "blue", lwd = 2)

# Add a legend with custom labels for each line
legend("topleft", legend = c("Rust", "R"), col = c("#D84B16", "blue"), lwd = 2)
```

While the example of the function I used is trivial, I can see other cases where it might be advantageous to migrate functions to Rust and then call those functions in R (better yet, just [do everything in Rust](https://www.nature.com/articles/d41586-020-03382-2) if possible). I could also try optimizing the R function via [memoization](https://en.wikipedia.org/wiki/Memoization) and then comparing its performance against its Rust counterpart. I leave that as an exercise for the reader.


